# **Preparing Date**  {-} 

## PM(2.5) vs Patient {-}

### Data Preparation {-}

Data set is from Synthetic patient and population health data from the state of Massachusetts


1.  Download complete patients FHIR file from <https://synthea.mitre.org/downloads> 22G zip tar

2.  Developed python [fhir_dataset.py](https://github.com/NIEHS/pcor_climate_tools/blob/main/geospatial_analysis/data_wrangling/fhir_dataset.py "File github location") to pull the patients information : patientid, lat, lon, birthdate, gender , start_date_time, end_date_time, code, code_display, start_date, end_date Total patients: 1 million patients

    sample result: ran synthea_1m_fhir_3\_0_May_24/output_12 code =183478001

    output file : patient_encounts_bycode_v2_12.csv

3.  QC dataset. for instance : start date \> end date ; misplace birth/death date to hospital stay date\

4.  Using the start/end date from patients file run DeGauss pm2.5 model <https://degauss.org/pm/>

    input file for pm2.5 from step 2
    command line:docker run --rm -v $PWD:/tmp ghcr.io/degauss-org/pm:0.2.0 tmp/yourdatafile.csv

5.  The following columns been added for each row:

    pm_pred : predicted PM2.5 (micrograms per cubic meter)

    pm_se : standard error for predicted PM2.5

    Output file: patient_encounts_bycode_v2_12_pm_0.2.0.csv

The output file will contain one row per day between start_date and end_date for everyone. Lat(latitude) and lon (longitude) location. This means that the output file will likely contain many more rows than the input file

6.  Run deGUAUSS census block group : <https://degauss.org/census_block_group/> Produced a file with additional columns:

    • census_block_group_id_2010: identifier for 2010 block group

    • census_tract_id_2010: identifier for 2010 tract

    Output file: patient_encounts_bycode_v2_12_pm_0.2.0_census_block_group_0.6.0_2010.csv
    rename to:ms_patient_pm_census_v2.csv

```{R}
library(readr)
df <- read_csv("./dataset/ms_patient_pm_census_v2.csv", show_col_types = FALSE)
```

```{r}
dim(df)
colnames(df)
```


install packages

```{r}
if (!require("shiny") ||
      !require("tidycensus") ||
      !require("tidyverse") ||
      !require("viridis")) {
  install.packages(c("shiny", "tidycensus", "tidyverse", "viridis"))
}
if (!require("plotly")) {
  install.packages("plotly", type = "source")
}
```

```{r }
if (!require("ggplot2")) {
  install.packages("ggplot2")
}
if (!require("maps") || !require("ggmap")) {
  install.packages(c("maps", "ggmap"))
}
```
Note that the echo = FALSE parameter was added to the code chunk to prevent printing of the R code that generated the plot. 

### Load libraries {#new-session}

```{r include=FALSE}
library(tidycensus)
library(tidyverse)
library(viridis)
library(shiny)
library(plotly)
# Only load the census key at the first time. then we set install=true
first <- FALSE
if (first == TRUE) {
  census_api_key(install = TRUE)
}
options(tigris_use_cache = TRUE)
```

Data wrangling

```{r}
# Check for missing values
sum(is.na(df))

# If there are missing values, you can drop or fill them as per your requirement
df <- na.omit(df) # Drop rows with missing values
```

Change column name for later map

```{r}
colnames(df)[which(names(df) == "lon")] <- "longitude"
colnames(df)[which(names(df) == "lat")] <- "latitude"
dim(df)
```

Filter the rows based on the date range Use the subset() function to filter the rows based on the date range:2010 -2017to match 5 year sensus data

```{r}
filtered_df <-
  subset(df, date >= as.Date("2012-01-01") & date <= as.Date("2017-09-01"))
dim(filtered_df)

write.csv(filtered_df, "./dataset/asthma_small.csv")
```

Combine information for display in tooltip

```{r}
# Combine the required information for hover tooltip into a new column
filtered_df$tooltip <- paste(
  "PM2.5:", filtered_df$pm_pred, "<br>",
  "Hospital Date:", filtered_df$date, "<br>",
  "Patient Info:", filtered_df$code_display, "<br>",
  "gender :", filtered_df$gender
)
```

### Create a Geographical Map

Load the required libraries for plotting maps:

```{r }
library(ggplot2)
library(maps)
library(ggmap)
```

Add map data to the base plot using the map_data() function:

```{R}
# Draw massachusetts map
ma_map <- map_data("state", region = "massachusetts")

# Create a base plot for Massachusetts
p <- ggplot() +
  geom_polygon(
    data = ma_map, aes(x = long, y = lat, group = group),
    fill = "lightgray", color = "black"
  )
```

```{R include=FALSE}
# print(p)
```

### Plotting the Data Points

Plot the data points on the map using the geom_point() function. Specify the longitude and latitude coordinates, and optionally, the color or size of the data points based on the PM2.5 values:

Add PM2.5 data points with hover information

```{r}
p <- p + geom_point(data = filtered_df, aes(
  x = longitude, y = latitude, color = pm_pred,
  text = tooltip
), size = 3)
# you can run this block code on your local to have mouse hover text.
# for Now it cannot
# rander on R-studio connect
# Convert the plot to a plotly object
run <- FALSE
if (run == TRUE) {
  p <- ggplotly(p) %>%
    layout(
      hoverlabel = list(bgcolor = "white"),
      hovertemplate = paste(
        "<b>%{text}</b>"
      )
    )
}
```

Draw the map

```{r}
print(p)
```


