### Visualize Air Pollution and Health Information Using Bivariate Maps {.unnumbered}

#### Set Up Bivariate Color Classes {-}

Each variate is divided into thirds (based on percentiles) and a joint classification for all 9 combinations is defined.

```{r}
require(latticeExtra) # uscancerrates, mapplot
require(maps) # map
require(classInt) # classIntervals, findCols
require(grid) # viewport, pushViewport
require(pals) # brewer.blues, stevens.pinkgreen
```

From asthma_patients_pm_2_census_block_group_0.6.0_2010.csv run python biv_cho_map.py call fcc api to get county FIPS and Name for each patient <https://geo.fcc.gov/api/census/#!/block/get_block_find> new data fileasthma_patients_pm_2_census_block_group_0.6.0_2010.csv

```{R}
library(readr)
df_county <- read.csv("./dataset/county.csv")
head(df_county)

df_census_block <- read.csv("./dataset/asthma_patients_pm_2_census_block_group_0.6.0_2010.csv")
head(df_census_block)
```

#### Get MA County Boundaries {-}

```{R include=FALSE}
library(sf)
library(ggplot2)
library(dplyr)
library(viridis)
library(tigris)

us_counties <- counties(cb = TRUE, year = 2020, class = "sf")

us_counties_MA <- us_counties %>%
  dplyr::filter(STATEFP == "25") # FIPS code for MA is 25
dim(us_counties_MA)
```

Aggregate the data by county, calculating the total number of patients and the mean pm2.5_pred:

```{R}
df_agg <- df_county %>%
  group_by(county_code) %>%
  reframe(
    patient_count = n(),
    county_name,
    pm2_mean = mean(pm_pred, na.rm = TRUE)
  )
```

### Join Census Data with County Boundaries

```{R}
dim(df_agg)
ma_counties_data <- us_counties_MA %>% left_join(df_agg, by = c("NAMELSAD" = "county_name"))
head(ma_counties_data)
```

```{r, eval = FALSE}
ggplot() +
  geom_sf(data = ma_counties_data,
          aes(fill = patient_count),
          color = NA) +
  scale_fill_viridis() +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(fill = "Patient count")
```

#### Set Color on Both Varaibles {-}

```{r}
cols <- stevens.pinkgreen
nbins <- 3

# categorize rates into 3 percentile bins
brks_pm <- classIntervals(
  log(ma_counties_data$pm2_mean),
  n = nbins,
  style = "quantile"
)
```

```{r}
brks_patient <- classIntervals(
  log(ma_counties_data$patient_count),
  n = nbins,
  style = "quantile"
)
```

```{r}
class_pm <- findCols(brks_pm)
print(class_pm)
```

```{R}
class_patient <- findCols(brks_patient)
# convert x,y classes into a joint class x+3(y-1)
ma_counties_data$class2 <- class_pm + nbins * (class_patient - 1)

# scatterplot of two variates showing bins
plot(log(pm2_mean) ~ log(patient_count),
  data = ma_counties_data,
  col = cols()[ma_counties_data$class2], pch = 19,
  xlim = c(0, 6), ylim = c(0, 5)
)
```

#### Map Air Pollution and Patient Counts {-}

```{r, eval = FALSE}
# Check for missing values
sum(is.na(ma_counties_data))

# If there are missing values, you can drop or fill them as per your requirement
df <- na.omit(ma_counties_data)  # Drop rows with missing values
```

```{r}
m4 <- mapplot(rownames(ma_counties_data) ~ class2,
  data = ma_counties_data,
  colramp = cols, breaks = seq(from = 0.5, by = 1, length = nbins * nbins + 1),
  xlab = "",
  colorkey = FALSE,
  map = map("state", "massachusetts",
    plot = FALSE, fill = TRUE,
    projection = "tetra"
  ),
  scales = list(draw = FALSE)
)
```

```{r, eval = FALSE}
print(nbins)
matrix(1:(nbins * nbins), nrow = nbins)
```

```{r}
m4leg <- levelplot(matrix(1:(nbins * nbins), nrow = nbins),
  axes = FALSE, col.regions = cols(),
  xlab = "patient", ylab = "pm_2.5", cuts = 8, colorkey = FALSE,
  scales = list(draw = 0)
)

vp <- viewport(x = .15, y = .25, width = .3, height = .3)
pushViewport(vp)
print(m4leg, newpage = FALSE)
popViewport()
```

```{r, eval = FALSE}
suppressWarnings(print(m4))

table(apply(data_sample, 1, paste, collapse = ""))
```

```{r}
m4leg <- levelplot(matrix(1:(nbins * nbins), nrow = nbins),
  axes = FALSE, col.regions = cols(),
  xlab = "patient", ylab = "pm_2.5", cuts = 8, colorkey = FALSE,
  scales = list(draw = 0)
)

# add the color legend
m4leg <- levelplot(matrix(1:(nbins * nbins), nrow = nbins),
  axes = FALSE, col.regions = cols(),
  xlab = "patient", ylab = "pm_2.5", cuts = 8, colorkey = FALSE,
  scales = list(draw = 0)
)
vp <- viewport(x = .15, y = .25, width = .3, height = .3)
pushViewport(vp)
print(m4leg, newpage = FALSE)
popViewport()
```
