### Mapping Linked Census Income Data 

#### Data Preparation {-}

Data set is from Synthetic patient and population health data from the state of Massachusetts

1.  Download complete patients FHIR file from <https://synthea.mitre.org/downloads> 22G zip tar

2.  Developed python [fhir_dataset.py](https://github.com/NIEHS/pcor_climate_tools/blob/main/geospatial_analysis/data_wrangling/fhir_dataset.py "File github location") to pull the patients information : patientid, lat, lon, birthdate, gender , start_date_time, end_date_time, code, code_display, start_date, end_date Total patients: 1 million patients

    sample result: ran synthea_1m_fhir_3\_0_May_24/output_12 code =183478001

    output file : patient_encounts_bycode_v2_12.csv

3.  QC dataset. for instance : start date \> end date ; misplace birth/death date to hospital stay date\

4.  Using the start/end date from patients file run DeGauss pm2.5 model <https://degauss.org/pm/>

    input file for pm2.5 from step 2 command line:docker run --rm -v \$PWD:/tmp ghcr.io/degauss-org/pm:0.2.0 tmp/yourdatafile.csv

5.  The following columns been added for each row:

    pm_pred : predicted PM2.5 (micrograms per cubic meter)

    pm_se : standard error for predicted PM2.5

    Output file: patient_encounts_bycode_v2_12_pm_0.2.0.csv

The output file will contain one row per day between start_date and end_date for everyone. Lat(latitude) and lon (longitude) location. This means that the output file will likely contain many more rows than the input file

6.  Run deGUAUSS census block group : <https://degauss.org/census_block_group/> Produced a file with additional columns:

    • census_block_group_id_2010: identifier for 2010 block group

    • census_tract_id_2010: identifier for 2010 tract

    • command line:docker run --rm -v \$PWD:/tmp ghcr.io/degauss-org/census_block_group:0.5.0 tmp/yourdatafile.csv

    Output file: yourdatafile_census_block_group_0.6.0_2010.csv rename to:ms_patient_pm_census_v2.csv

install packages

```{r}
if (!require("shiny") ||
      !require("tidycensus") ||
      !require("tidyverse") ||
      !require("viridis")) {
  install.packages(c("shiny", "tidycensus", "tidyverse", "viridis"))
}
if (!require("plotly")) {
  install.packages("plotly", type = "source")
}
```

```{r }
if (!require("ggplot2")) {
  install.packages("ggplot2")
  install.packages(c("maps", "ggmap"))
}
if (!require("maps") || !require("ggmap")) {
  install.packages(c("maps", "ggmap"))
}
```

Note that the echo = FALSE parameter was added to the code chunk to prevent printing of the R code that generated the plot. 
### Load libraries

```{r include=FALSE}
library(tidycensus)
library(tidyverse)
library(viridis)
library(shiny)
library(plotly)
# Only load the census key at the first time. then we set install=true
first <- FALSE
if (first == TRUE) {
  census_api_key(install = TRUE)
}
options(tigris_use_cache = TRUE)
```

```{R}
library(readr)
df <- read_csv("./dataset/ms_patient_pm_census_v2.csv", show_col_types = FALSE)
```

```{r}
dim(df)
colnames(df)
```

#### Get Census Income Data {-}

7.  Apply census api key at <https://api.census.gov/data/key_signup.html> and then supply the key to the `census_api_key()` function to use it throughout your tidycensus session.

8.  Run r package tidycensus with the api key and the parameters:

    • income code = "C17002_001","C17002_002","C17002_003","C17002_004",

    "C17002_005","C17002_006","C17002_007", "C17002_008"

    • geography = tract • year = 2012 • state = MA • survey = acs5

Create a character vector named income_code containing eight elements. Each element represents a variable code related to income.

```{R}
income_code <- c(
  "C17002_001", "C17002_002", "C17002_003", "C17002_004", "C17002_005", "C17002_006",
  "C17002_007", "C17002_008"
)
# Assign the result of the get_acs function to the variable tarr. The function retrieves
# American Community Survey (ACS) data for the     specified geography (tract), variables
# (the income_code vector), state (Massachusetts with state code 25), geometry (TRUE to
# include spatial information), survey ("acs5"), and year (2012).

tarr <- get_acs(
  geography = "tract", variables = income_code,
  state = 25, geometry = FALSE, survey = "acs5", year = 2012
)

# Write the census income data into csv file
# modify if (FALSE) to if (TRUE) if you run the first time
if (FALSE) {
  write.csv(tarr, "./dataset/income_MA_mult_incomes.csv")
}
```

9.  tarr column name : • Id -- census track id • geoid, • name, • variable, • estimate, • moe

10. Join the patient's information + degauss pm 2.5 + census block dataset with census income dataset by census track id

11. Data join output file : asthma_patients_pm_2\_census_block_group_0.6.0_2010.csv

```{r}
dim(tarr)
colnames(tarr)
```

#### Link Census Income Data by Tract Geoid {-}

df column "census_block_group_id_2010" and tarr column "GEOID" are the census track id.

Join these two data frames

```{r}
library(dplyr)
merge <- FALSE
if (merge == TRUE) {
  merged_frame <- df %>%
    mutate(census_tract_id_2010 = as.character(census_tract_id_2010)) %>%
    inner_join(tarr, by = c("GEOID" = "census_tract_id_2010"))
}

# left join produce the same result as inner_join
left_merged_frame <- df %>%
  mutate(census_tract_id_2010 = as.character(census_tract_id_2010)) %>%
  left_join(tarr, by = c("census_tract_id_2010" = "GEOID"))

# Write the merged frame data into csv file
# modify if (FALSE) to if (TRUE) if you run the first time
if (FALSE) {
  write.csv(left_merged_frame, "./dataset/merged_income_MA_mult_incomes.csv")
}
```

```{r}
library(dplyr)
```

#### Identify County with Hightest Income in 2020 {-}

```{r}
# Fetch ACS data for Massachusetts counties
ma_counties <- get_acs(
  geography = "county",
  variables = "B19013_001",
  state = "MA",
  year = 2020
)

# Sort the data by median household income in descending order
ma_counties_sorted <- ma_counties %>%
  arrange(desc(estimate))

# Extract the county with the highest income
highest_income_county <- ma_counties_sorted$NAME[1]

# Print the county with the highest income
cat("The county with the highest income in Massachusetts is:",
    highest_income_county)
```

#### Map Income and Patient Data {-}

```{r}
if (!require("maps") || !require("ggmap")) {
  install.packages(c("maps", "ggmap"))
  library(maps)
  library(ggmap)
}
library(ggplot2)
```
```{r}
colnames(left_merged_frame)
```

```{r}
# Draw Massachusetts map
ma_map <- map_data("state", region = "massachusetts")
# modify column's name
colnames(left_merged_frame)[which(names(left_merged_frame) == "lon")] <-
  "longitude"
colnames(left_merged_frame)[which(names(left_merged_frame) == "lat")] <-
  "latitude"

# Optional Filter the rows based on the date range Use the subset() function to
# filter the rows based on the date range:2012 -2017to match 5 year sensus data

filtered_df <- subset(
  left_merged_frame,
  date >= as.Date("2012-09-01") & date <= as.Date("2017-09-01")
)

# Create a base plot for Massachusetts
p <- ggplot() +
  geom_polygon(
    data = ma_map, aes(x = long, y = lat, group = group),
    fill = "lightgray", color = "black"
  ) +
  # Plot the data points
  geom_point(
    data = filtered_df,
    aes(x = longitude, y = latitude, color = estimate),
    size = 3
  ) +
  labs(title = "Income vs patient Map", x = "Longitude", y = "Latitude") +
  scale_color_gradient(low = "blue", high = "red", name = "estimate") +
  theme_minimal() +
  # Adjust the map boundaries to focus on Massachusetts
  coord_cartesian(xlim = c(-73.5, -69.9), ylim = c(41.2, 42.9))

# Print the map
print(p)
```
